Stochastic Properties of EIP-1559 Basefees

### 2021-05-14 eip1559/template.tex
paper corrections

\documentclass[peerreview]{ieeesyscoin}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage[switch]{lineno}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}
\linenumbers
\history{}

\title{\centering Stochastic Properties of EIP-1559 Basefees}
\author{\centering  \uppercase{Ian C. Moore, PhD}\authorrefmark{1}, 
and \uppercase{Jagdeep Sidhu, MSc\authorrefmark{2}}}

\address[1]{\centering  (e-mail: ic3moore@gmail.com)}
\tfootnote{}
\address[2]{\centering Syscoin Core Developer, Blockchain Foundry Inc.(e-mail: jsidhu@blockchainfoundry.co)}

\markboth
{Sidhu \headeretal: Stochastic Properties of EIP-1559 Basefees}
{Sidhu \headeretal: Stochastic Properties of EIP-1559 Basefees}

\corresp{}

\begin{abstract}
EIP-1559 is a new proposed pricing mechanism for the Ethereum protocol developed to bring stability to fluctuating gas prices. To properly understand this as a stochastic process, it is necessary to develop the mathematical foundations to understand under what conditions does the base fee gas price outcomes behave as a stationary process, and when it does not. Understanding these mathematical fundamentals is critical to properly engineering a stable system.
\end{abstract}

\begin{keywords}
EIP 1559, Base Fees, Stochastic Processes, Stationarity
\end{keywords}

\titlepgskip=-15pt

\maketitle

\section{Introduction}
\label{sec:introduction}

The current Ethereum pricing mechanism employs the price auction model where high-value use cases are prioritized over lower ones. The problem with this approach is that as Ethereum has grown more popular, users have difficulty estimating optimal gas fees. In 2018, EIP-1559 was  proposed by Ethereum Founder, Vitalik Buterin, as a major change to Ethereum's transaction fee mechanism, and introduces two different types of fees, which is the base fee and inclusion fee. The ideas is the base fee is relatively stable while the inclusion fee serves as an additional tip offered to compensate miners. This mechanism allows for the base fee to be adjusted after every block in accordance to demand. Hence, when demand for gas is higher than the target gas price, the block size is adjusted upwards, and when demand for gas is lower than the target gas, the block size is adjusted downwards. Blocks are allowed to grow as large as double the target block size.

@@

\begin{equation}
b_{k} = b_{k-1}f_{k-1}
\label{eq:eip1559} \tag{1}
\end{equation}

\begin{equation}
f_{k} = 1 + \frac{\delta_{k}}{c}
\label{eq:fk} \tag{2}
\end{equation}
where $\delta_{k} = demand_{k} - Target Gas Fee$ and $c = Target ~/~ BaseFee Max Change$; for the purpose of this study, we assume $\delta_{k}$ to be a random stationary process. However before we do that, we run abm1559 simulator to show that this is the case, under the assumption that new users are Poisson distributed.

@@

 \begin{equation}
(X_{1} , ... , X_{N} )^T \stackrel{D}{=} (X_{1+\tau} , ... , X_{N+\tau} )^T,
\label{eq:ar1} \tag{3}
\end{equation}
where $\stackrel{D}{=}$ means equal in distribution.

@@

\begin{equation}
\delta y_{t} = \alpha y_{t-1} + \theta_{1} \delta Y_{t-1}  + ... + \theta_{p} \delta Y_{t-p} + \epsilon_{t},
\label{eq:ar1} \tag{4}
\end{equation}
the Augmented Dickey Fuller (ADF) test checks the existence of a unit root $\alpha = 1$ (ie, H0: $\alpha = 1$) where the null hypothesis is non-stationary. The unit root in (\ref{eq:ar1}) is characteristic of a time series that makes it non-stationary.

@@

To simulate gas demand $\delta_{k}$ in (\ref{eq:fk}) we use the abm1559 simulator \cite{Mon21} where the following assumptions were made: (a) users are Poisson distributed; and (b) every TX consumes 21000 Gas. For our analysis, we are interested in simulating a year's worth of gas costs, which is infeasible as running the full Ethereum chain simulator consumes a lot of overhead.

@@

\begin{figure}[h!]
\includegraphics[width=3in]{img/gas.png}
\caption{Simulations of; (top) gas demands using random samples from a normal distribution; and (bottom) gas costs using (\ref{eq:eip1559}). As we can see, the gas cost simulations (bottom) are behaving like a random walk process} 
\label{fig:gas}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=3in]{img/gaussian_analysis.png}
\caption{Testing normality of gas demands (top image of Fig. \ref{fig:gas}) using; (left) histogram with normal distribution fit of demands; and (right) Normal Q-Q plot indicates sample and theoretical quantiles match} 
\label{fig:eda}
\end{figure}

We also applied the ADF test to the simulated gas demands, shown in top image of Figure \ref{fig:gas}, and found it to be statistically significant (ie, p-val = 2.33e-29). Hence, rejecting HO indicating simulated gas demands to be stationary.

@@

In Section \ref{section:gas_demand} we simulated gas demand  $\delta_{k}$ in (\ref{eq:fk}) and found them to be stationary and normally distributed with $\mu$ = 13641075, and variance $\sigma^2$ = 550980. To simulate gas prices using EIP 1559 we used these parameter estimates to model out a year's worth of gas demands and fed it through the system described in (\ref{eq:eip1559}) and (\ref{eq:fk}). A series of these simulations can be seen in Figure \ref{fig:basefee_simulations}.

\begin{figure}[h!]
\includegraphics[width=3in]{img/basefee_simulations.png}
\caption{Gas cost simulations using EIP 1559; we can visual see the non-stationary behaviour in price over time, which is indicative of a random walk.}  
\label{fig:basefee_simulations}
\end{figure}

@@

\begin{table}[h!]
\centering
\begin{tabular}{ |c|c|c|c| } 
\hline
 Simulation & ADF Statistic & p-value & Outcome \\
\hline
1 & -1.506 & 0.531 & not stationary \\
2 & -1.260 & 0.647 & not stationary \\
3 & -2.170 & 0.217 & not stationary \\
4 & -2.432 & 0.133 & not stationary \\
5 & -1.844 & 0.359 & not stationary \\
6 & -1.323 & 0.618 & not stationary \\
7 & -1.645 & 0.459 & not stationary \\
8 & -1.662 & 0.451 & not stationary \\
9 & -1.124 & 0.705 & not stationary \\
10 & -1.930 & 0.318 & not stationary \\
\hline
\end{tabular}
\caption{ADF test on basefee simulations from Figure \ref{fig:basefee_simulations} to test null hypothesis (H0) that a unit root is present; when unit root is present, then sample is considered to be non-stationary. As we can see, we fail to reject H0 for all ten tests indicating insufficient evidence to conclude that the effect of stationarity exists.}
\label{table:pow_vs_pos}
\end{table}

@@

\begin{equation}
x_{t} = \mu + \alpha (x_{t-1} -\mu) + \epsilon_{t-1} ~~~~~~~~ \epsilon_{t} \sim IID~N(0,\sigma^2)
\label{eq:ar1} \tag{5}
\end{equation}
The above process is stationary when $|\alpha|$<1 so that this has a single root 1/$\alpha$ outside the unit circle.  It can be shown that if $\sigma$>0, then this process has the asymptotic stationary distribution:

\begin{equation}
x_{\infty} \sim N(\mu,\frac{\sigma^2}{1-\alpha^2})
\label{eq:x_infinity} \tag{6}
\end{equation}
Considering the (state space like) simularities to the AR(1) process to the system of the system of (\ref{eq:eip1559}) and (\ref{eq:fk}); the question is what are the constraints (if any) that time varying process $\delta_{t}$ must meet to ensure stationary outcomes for the basefees.

@@

\begin{equation}
x_{k} = \alpha + \beta_{k}x_{k-1} + \epsilon_{k},
\label{eq:rca1} \tag{7}
\end{equation}
where ${\epsilon_{k}}$ is an independent sequence of random variables with 0 mean and variance $\sigma^2 > 0$; ${\beta_{k}}$ is an independent sequence of random variables with mean $\mu_{\beta}$ and variance $\sigma_{\beta}^2$ of the random coefficient $\beta_{k}$, and variance $\sigma^2$ of the error $\epsilon_{k}$. Given (3) and (4) it is obvious that when $\sigma_{\beta}^2 = 0$, the RCA(1) in (5) becomes a AR series, and becomes a random walk when $\mu_{\beta} = 1$ and $\sigma_{\beta}^2 = 0$, hence non-stationary.

@@

\begin{equation}
\ln(\sigma_{\beta}^2) < \varsigma + \ln(2) - 2 \int_{0}^{1}\frac{1 - \exp[-\lambda(1-w^2)]}{1-w^2}dw,
\label{eq:thm} \tag{8}
\end{equation}
where $\varsigma$ $\approx$ 0.57721, denotes Euler’s constant and $\lambda$ = $\mu_{\beta}^2$/ $2\sigma_{\beta}^2$

\textbf{Proof:} See Theorem 2 in [1]

\begin{figure}[h!]
\includegraphics[width=3in]{img/strict_stationarity.png}
\caption{Region of strict stationarity and ergodicity for RCA(1) determined via (\ref{eq:thm}). The annotated red dot represents where our setup is in relation to the region where an RCA(1) process behaves as a strictly stationary process. Based on this we can see that our setup was non-stationary.} 
\label{fig:strict_stationarity}
\end{figure}

@@

\begin{thebibliography}{00}

@@

\bibitem{Wan03} Dazhe Wang, Frequentist and Bayesian Analysis of Random Coefficient Autoregressive Models, North Carolina State University PhD Dissertation, 2003

@@

\EOD

>>>>

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}		% Can be removed after putting your text content
\usepackage{graphicx}
\usepackage{doi}

\title{Stochastic Properties of EIP-1559 Basefees}

%\date{September 9, 1985}	% Here you can change the date presented in the paper title
%\date{} 					% Or removing it

\author{ \href{https://orcid.org/0000-0000-0000-0000}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Ian C. Moore, PhD}\\
	\texttt{ic3moore@gmail.com} \\
	%% examples of more authortexmaker
	\And
	\href{https://orcid.org/0000-0000-0000-0000}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Jagdeep Sidhu, MSc} \\
	Syscoin Core Developer\\
	Blockchain Foundry Inc.\\
	\texttt{jsidhu@blockchainfoundry.co} \\
	%% \AND
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
	%% \And
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
	%% \And
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
}

% Uncomment to remove the date
%\date{}

% Uncomment to override  the `A preprint' in the header
%\renewcommand{\headeright}{Technical Report}
%\renewcommand{\undertitle}{Technical Report}
\renewcommand{\shorttitle}{EIP-1559 Basefees}

%%% Add PDF metadata to help others organize their library
%%% Once the PDF is generated, you can check the metadata with
%%% $ pdfinfo template.pdf
\hypersetup{
pdftitle={Stochastic Properties of EIP-1559 Basefees},
pdfsubject={q-bio.NC, q-bio.QM},
pdfauthor={Ian C. Moore, Jagdeep Sidhu},
pdfkeywords={EIP 1559, Base Fees, Stochastic Processes, Stationarity},
}

\begin{document}
\maketitle

\begin{abstract}
EIP-1559 is a new proposed pricing mechanism for the Ethereum protocol developed to bring stability to fluctuating gas prices. To properly understand this as a stochastic process, it is necessary to develop the mathematical foundations to understand under what conditions the base fee gas price outcomes behave as a stationary process, and when it does not. Understanding these mathematical fundamentals is critical to properly engineering a stable system.
\end{abstract}

% keywords can be removed
\keywords{EIP 1559  \and Base Fees  \and Stochastic Processes  \and Stationarity}

\section{Introduction}
The current Ethereum pricing mechanism employs the price auction model where high-value use cases are prioritized over lower ones. The problem with this approach is that as Ethereum has grown more popular, users have difficulty estimating optimal gas fees. In 2018, EIP-1559 was  proposed by Ethereum Founder, Vitalik Buterin, as a major change to Ethereum's transaction fee mechanism, and introduces two different types of fees, which is the base fee and inclusion fee \cite{But19}. The ideas is the base fee is relatively stable while the inclusion fee serves as an additional tip offered to compensate miners. This mechanism allows for the base fee to be adjusted after every block in accordance to demand. Hence, when demand for gas is higher than the target gas price, the block size is adjusted upwards, and when demand for gas is lower than the target gas, the block size is adjusted downwards. Blocks are allowed to grow as large as double the target block size.

@@

\begin{equation}
b_{k} = b_{k-1}f_{k}
\label{eq:eip1559} 
\end{equation}

\begin{equation}
f_{k} = 1 + \frac{\delta_{k}}{c}
\label{eq:fk} 
\end{equation}
where $\delta_{k} = demand_{k} - Target Gas Fee$ and $c = Target ~/~ BaseFee Max Change$; for the purpose of this study, we assume $\delta_{k}$ to be a random stationary process. However before we do that, we run the \textit{abm1559} simulator \cite{Mon21} to show that this is the case, under the assumption that new users are Poisson distributed.

@@

 \begin{equation}
(X_{1} , ... , X_{N} )^T \stackrel{D}{=} (X_{1+\tau} , ... , X_{N+\tau} )^T,
\label{eq:ar1} 
\end{equation}
where $\stackrel{D}{=}$ means equal in distribution.

@@

\begin{equation}
\delta y_{t} = \alpha y_{t-1} + \theta_{1} \delta Y_{t-1}  + ... + \theta_{p} \delta Y_{t-p} + \epsilon_{t},
\label{eq:ar1} 
\end{equation}
the Augmented Dickey Fuller (ADF) test checks the existence of a unit root $\alpha = 1$ (ie, H0: $\alpha = 1$) where the null hypothesis is non-stationary. The unit root in (\ref{eq:ar1}) is characteristic of a time series that makes it non-stationary.

@@

To simulate gas demand $\delta_{k}$ in (\ref{eq:fk}) we use the \textit{abm1559}  simulator \cite{Mon21} where the following assumptions were made: (a) users are Poisson distributed; and (b) every TX consumes 21000 Gas. For our analysis, we are interested in simulating a year's worth of gas costs, which is infeasible as running the full Ethereum chain simulator consumes a lot of overhead.

@@

\begin{figure}
\centering
\includegraphics[width=5in]{gas.png}
\caption{Simulations of; (top) gas demands using random samples from a normal distribution; and (bottom) gas costs using (\ref{eq:eip1559}). As we can see, the gas cost simulations (bottom) are behaving like a random walk process} 
\label{fig:gas}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=5in]{gaussian_analysis.png}
\caption{Testing normality of gas demands (top image of Fig. \ref{fig:gas}) using; (left) histogram with normal distribution fit of demands; and (right) Normal Q-Q plot indicates sample and theoretical quantiles match} 
\label{fig:eda}
\end{figure}

We also applied the ADF test to the simulated gas demands, shown in top image of Figure \ref{fig:gas}, and found it to be statistically significant (ie, p-val = 2.33e-29). Hence, rejecting HO, which indicate simulated gas demands to be stationary.

@@

In Section \ref{section:gas_demand} we simulated gas demand  $\delta_{k}$ in (\ref{eq:fk}) and found them to be stationary and normally distributed with $\mu$ = 13641075, and variance $\sigma^2$ = 550980. To simulate gas prices using EIP 1559 we used these parameter estimates to model a year's worth of gas demands and fed it through the system described in (\ref{eq:eip1559}) and (\ref{eq:fk}). A series of these simulations can be seen in Figure \ref{fig:basefee_simulations}.

\begin{figure}
\centering
\includegraphics[width=5in]{basefee_simulations.png}
\caption{Gas cost simulations using EIP 1559; we can visual see the non-stationary behaviour in price over time, which is indicative of a random walk.}  
\label{fig:basefee_simulations}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{ |c|c|c|c| } 
\hline
 Simulation & ADF Statistic & p-value & Outcome \\
\hline
1 & -1.506 & 0.531 & not stationary \\
2 & -1.260 & 0.647 & not stationary \\
3 & -2.170 & 0.217 & not stationary \\
4 & -2.432 & 0.133 & not stationary \\
5 & -1.844 & 0.359 & not stationary \\
6 & -1.323 & 0.618 & not stationary \\
7 & -1.645 & 0.459 & not stationary \\
8 & -1.662 & 0.451 & not stationary \\
9 & -1.124 & 0.705 & not stationary \\
10 & -1.930 & 0.318 & not stationary \\
\hline
\end{tabular}
\caption{ADF test on basefee simulations from Figure \ref{fig:basefee_simulations} to test null hypothesis (H0) that a unit root is present; when unit root is present, then sample is considered to be non-stationary. As we can see, we fail to reject H0 for all ten tests indicating insufficient evidence to conclude that the effect of stationarity exists.}
\label{table:pow_vs_pos}
\end{table}

@@

\begin{equation}
x_{t} = \mu + \alpha (x_{t-1} -\mu) + \epsilon_{t-1} ~~~~~~~~ \epsilon_{t} \sim IID~N(0,\sigma^2)
\label{eq:ar1} 
\end{equation}
The above process is stationary when $|\alpha|$<1 so that this has a single root 1/$\alpha$ outside the unit circle.  It can be shown that if $\sigma$>0, then this process has the asymptotic stationary distribution:

\begin{equation}
x_{\infty} \sim N(\mu,\frac{\sigma^2}{1-\alpha^2})
\label{eq:x_infinity} 
\end{equation}
Considering the (state space like) simularities to the AR(1) process to the system of the system of (\ref{eq:eip1559}) and (\ref{eq:fk}); the question is what are the constraints (if any) that time varying process $\delta_{t}$ must meet to ensure stationary outcomes for the basefees.

@@

\begin{equation}
x_{k} = \alpha + \beta_{k}x_{k-1} + \epsilon_{k},
\label{eq:rca1}
\end{equation}
where ${\epsilon_{k}}$ is an independent sequence of random variables with 0 mean and variance $\sigma^2 > 0$; ${\beta_{k}}$ is an independent sequence of random variables with mean $\mu_{\beta}$ and variance $\sigma_{\beta}^2$ of the random coefficient $\beta_{k}$, and variance $\sigma^2$ of the error $\epsilon_{k}$. Given (3) and (4) it is obvious that when $\sigma_{\beta}^2 = 0$, the RCA(1) in (5) becomes a AR series, and becomes a random walk when $\mu_{\beta} = 1$ and $\sigma_{\beta}^2 = 0$, hence non-stationary.

@@

\begin{equation}
\ln(\sigma_{\beta}^2) < \varsigma + \ln(2) - 2 \int_{0}^{1}\frac{1 - \exp[-\lambda(1-w^2)]}{1-w^2}dw,
\label{eq:thm} 
\end{equation}
where $\varsigma$ $\approx$ 0.57721, denotes Euler’s constant and $\lambda$ = $\mu_{\beta}^2$/ $2\sigma_{\beta}^2$

\textbf{Proof:} See Theorem 2 in \cite{Wan03}

\begin{figure}
\centering
\includegraphics[width=5in]{strict_stationarity.png}
\caption{Region of strict stationarity and ergodicity for RCA(1) determined via (\ref{eq:thm}). The annotated red dot represents where our setup is in relation to the region where an RCA(1) process behaves as a strictly stationary process. Based on this we can see that our setup was non-stationary.} 
\label{fig:strict_stationarity}
\end{figure}

@@

\begin{thebibliography}{1}

\bibitem{But19} V. Buterin,  \textit{Blockchain Resource Pricing}, Apr 2019. Accessed on: May 2021. [Online]. Available: https://github.com/ethereum/research/blob/master/papers/pricing/ethpricing.pdf

@@

\bibitem{Wan03} D. Wang, Frequentist and Bayesian Analysis of Random Coefficient Autoregressive Models, North Carolina State University PhD Dissertation, 2003

@@
